---
title: "MarchMadness"
author: "Nathan Snyder.1896"
date: "2025-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)

```

```{r}
UpsetSeedInfo = read.csv("Upset Seed Info.csv")
EvanMiya <- read.csv("EvanMiya.csv")
TeamResults <- read.csv("Team Results.csv")
KenPom <- read.csv("KenPom Barttorvik.csv")
```


```{r}
#number of upets per year with difference 5 or more

numBigUpsetsPerYr <- UpsetSeedInfo %>% 
  filter(SEED.DIFF >= 5) %>%  
  group_by(YEAR) %>%       
  summarise(upset_count = n())

ggplot(numBigUpsetsPerYr, 
       aes(x = YEAR,
           y = upset_count)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = upset_count), vjust = -0.3) +
  labs(
    title = "Number of upsets with seed diff >= 5 by year",
    x = "Year",
    y = "Number of upsets with seed diff >= 5 "
  )

```
^(not used in presentation)

```{r}
#look at most common differential for upsets

diffCount <- UpsetSeedInfo%>%  
  group_by(SEED.DIFF) %>%       
  summarise(upset_count = n())

upsetbar = ggplot(diffCount, 
       aes(x = SEED.DIFF,
           y = upset_count)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(
    breaks = diffCount$SEED.DIFF
  ) +
  labs(
    title = "Number of Upsets All Time by Seed Difference",
    x = "Seed Difference",
    y = "Number of Upsets"
  ) +
  theme_minimal()

ggsave("upset_barplot.png", plot = upsetbar, width = 8, height = 5, dpi = 300)

```


```{r}
upset_counts <- UpsetSeedInfo %>%
  group_by(YEAR, SEED.DIFF) %>%
  summarise(upset_count = n(), .groups = "drop")

upsetheatmap = ggplot(upset_counts, aes(x = factor(YEAR), y = SEED.DIFF, fill = upset_count)) +
  geom_tile() +
  scale_fill_gradient(low = "gray90", high = "red") +   
  labs(
    x = "Year",
    y = "Seed Difference",
    fill = "Upset Count",
    title = "Upsets by Seed Difference & Year"
  ) +
  theme_minimal() 

ggsave("upsetheat.png", plot = upsetheatmap, width = 8, height = 5, dpi = 300)
```





Look at team's preseason vs actual success


Not used in presentation v
```{r}
KenPomPreseason <- read.csv("KenPom Preseason.csv")

KenPomAll <- KenPom %>% 
  inner_join(KenPomPreseason, by = c("TEAM", "YEAR", "TEAM.NO"))

ggplot(KenPomAll, aes(x = PRESEASON.KADJ.EM, y = KADJ.EM)) +
  geom_point(color = "blue", alpha = 0.7) +  
  labs(title = "Preseason EM vs Actual EM",
       x = "Preseason EM",
       y = "Actual EM") +
  theme_minimal()

```



```{r}
#try regular EM - preseason EM?
#nevermind this already exists - EM.change

top10EMChange <- KenPomAll %>% 
  arrange(desc(KADJ.EM.CHANGE)) %>% 
  head(10) %>% 
  select(TEAM, KADJ.EM.CHANGE)


ggplot(top10EMChange, 
       aes(x = reorder(as.factor(TEAM), -KADJ.EM.CHANGE),
           y = KADJ.EM.CHANGE)) +
  geom_col(fill = "steelblue") + 
  labs(
    title = "Top 10 teams in EM increase",
    x = "Team",
    y = "EM increase"
  )

```
(not used in presentation)


now show how this correlates to success

```{r}
#add column for "games won" - for each round, games won increases

KenPomAll <- KenPomAll %>%
  mutate(GamesWon = case_when(
    ROUND.x == 68 | ROUND.x == 64 ~ "0",
    ROUND.x == 32 ~ "1",
    ROUND.x == 16 ~ "2",
    ROUND.x == 8 ~ "3",
    ROUND.x == 4 ~ "4",
    ROUND.x == 2 ~ "5",
    ROUND.x == 1 ~ "6"
  ))


EMboxplot <- ggplot(KenPomAll, aes(x = as.factor(GamesWon), y = KADJ.EM.CHANGE)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "EM Change by Number of Games Won in Tournament",
       x = "Games Won",
       y = "EM Change") +
  theme_minimal()

ggsave("EMboxplot.png", plot = EMboxplot, width = 8, height = 5, dpi = 300)


```


```{r}
EMscatter = ggplot(KenPomAll, aes(x = GamesWon, y = KADJ.EM.CHANGE)) +
  geom_point(color = "blue", alpha = 0.7) +  
  labs(title = "Number of Games Won in Tournament vs EM Change",
       x = "Games Won",
       y = "EM Change") +
  theme_minimal()

ggsave("EMscatter.png", plot = EMscatter, width = 8, height = 5, dpi = 300)


```
There's obviously more data for the earlier rounds since less teams make it far, but still enough since the data is 2008-2024 total. It looks like there may be a correlation between more games won and a higher EM Change (increase). That is, teams that improve from preseason to regular season seem to have a better chance of making it farther in the tournament.



Pace
```{r}

MiyaPom <- EvanMiya %>% 
  inner_join(KenPomAll, by = c("TEAM", "YEAR"))

```



```{r}
#make scatter plot with win percentage vs pace adjust rank
#^that would only work with regular season

#you could also just see how many tournament games each team won each year


PaceWinPct = ggplot(MiyaPom, aes(x = WIN., y = PACE.ADJUST)) +
  geom_point(color = "blue", alpha = 0.7) +  # Scatter points
  labs(title = "Pace Adjusted vs. Win Percentage (Regular Season)",
       x = "Regular Season Win Percentage",
       y = "Pace Adjusted") +
  theme_minimal()

ggsave("PaceWinPct.png", plot = PaceWinPct, width = 8, height = 5, dpi = 300)
```

```{r}
MiyaPomNo2025 <- MiyaPom %>% 
  filter(YEAR != 2025)

PacevsGamesWon <- ggplot(MiyaPomNo2025, aes(x = GamesWon, y = `PACE.ADJUST`)) +
  geom_jitter(color = "blue", alpha = 0.7) +  # Scatter points
  labs(title = "Pace Adjusted vs. Games Won in Tournament",
       x = "Games Won",
       y = "Pace Adjusted") +
  theme_minimal()

ggsave("PaceGamesScatter.png", plot = PacevsGamesWon, width = 8, height = 5, dpi = 300)
```





Not used in slides v
```{r}
ggplot(MiyaPomNo2025, aes(x = GamesWon, y = KADJ.T)) +
  geom_jitter(color = "blue", alpha = 0.7) +  # Scatter points
  labs(title = "Tempo vs. Games Won in Tournament",
       x = "Games Won",
       y = "Tempo") +
  theme_minimal()
```

